Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-06-11 Manuscript/proto-manuscript.tex
Updates

@@

Simulations with molecular mechanics models represent nuclei as classical particles 
and describe their interactions through a pariwise additive empirical potential \cite{SOME_review, FF_papers}. 
According to the electronic continuum correction (ECC), electrons that surround 
the nuclei shall be taken into account in such simulations as an electronic continuum. \cite{leontyev11}
Such a continuum responds to the electrostatic interactions of the explicitly represented nuclei
as a polarizable dielectric. 
The dielectric constant of shuch a dielectric medium, $\epsilon _{el}$, will 
then resduce the electrostatic interactions in between the nuclei. 
This can be formally represented in classical simulations through scaling of charges of particles with a factor $f_q = \epsilon _{el} ^{-1/2}$. 
Note that the high frequency dielectric constant, which is experimentally measurable \cite{some_original_work, leontyev11}, represents
the part of dielectric response only due to electrons, and it is approximately $\epsilon _{el} \approx 2$ for almost any biomaterial that exists in living cells. 
The high frequency dielectric constant of water, 
the most abundant molecule in any cell, 
is $\epsilon _{el} = 1.78$, hence the scaling factor adopts the value of $f_q = 0.75$. \cite{some_orig_source, leontyev11}

The newly presented ECC-POPC model is based on the POPC model from Lipid14~\cite{dickson14} force field. 
There, the partial atomic charges were derived by 
fitting the electrostatic potential of its model quantum chemistry representation in vacuum. 
If such charges are obtained in an implicit solvent, they vary the most for the polar moieties \cite{maiejewski14}. 
In order to represent this difference in polarity in a model with fixed charges, 
we shall use the average charges from both environments (IpolQ charges) \cite{Cerutti2013}. 
By taking the charges of oxygen atoms in vacuum and implicit water solvent for POPC from \cite{maciejewski14}, 
we represent the IpolQ charges in the electronic continuum correction of the Lipid14 model
by increasing the scaling factor for charges to $f_q = 0.8$.

Because the biologically relevant cations like \ce{Na^+} and \ce{Ca^{2+}}
are found to interact only with the phospholipid head group,
and the accuracy of alkyl sidechains in Lipid14 model are satisfactory, 
we exclude alkyl tails from our modifications.

By scaling the particle charges alone, we reduce the 
electrostatic interactions and, hence, their contribution to hydration.
A simple approach to correct for this effect is to scale also 
the sigma factors in the Van der Waals interaction that is 
modelled with a Lennard-Jones potential. \cite{lennard-jones_work}
In line with the electronic continnum correction,
the scaling factor for sigma parameters, $f_\sigma$,
has to lay between $f_q$ and 1.

The optimal value for $f_\sigma$ was found by representing the overall membrane structure well
by matching scattering form factors to experimental data~\cite{Petrache06, Kucerka08, Pabst10}.
This ensures that the presented model describes correctly the membrane strucutre,
adopts a liquid disordered phase, 
and provides area per lipid agreeing with experiments (see Fig~\ref{simVSexpNOions} and Table~\ref{tab:apls}).

Detailed and robust structural information of lipid bilayers can
be reached with C-H bond order parameter from NMR experiments,
which tell on the lipid structure sampled by each individual molecule.
On the other hand, form factors from scattering experiments validate overall bilayer properties, 
lateral density (area per molecule) and thickness \cite{ollila16}. 
The structural quality of
new ECC lipid model is evaluated against scattering form factors and NMR order parameters
in Fig. \ref{simVSexpNOions}
\todo{Add acyl chain order parameters, POPC chemical structure}
\todo{Why original Lipid14 model seems to deviate from experiments, in contrast to the original publication? Different simulation setup used?}
\begin{figure}[]
  \centering
  \includegraphics[height=8.5cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
%  \caption{\label{fig:form-f}
%    X-ray scattering form factors from experiments~\cite{Kucerka2011} and simulations using Lipid14 and ECCLipids17 models.  }
%\end{figure}
%
%\begin{figure}[]
%  \centering
  \includegraphics[width=9.0cm]{../Fig/ipython_nb/Headgr_OPs_exp-L14-ECCL17.png}
  \caption{\label{simVSexpNOions}
    X-ray scattering form factors from experiments~\cite{Kucerka2011} and simulations using Lipid14 and ECCLipids17 models. 
    Headgroup and glycerol backbone order parameters from simulations with Lipid14 \cite{dickson14} and EECLipid17 models
    compared with experimental order parameters from \cite{ferreira13}.}
\end{figure}
Area per molecules extracted from MD simulations and SPD model fitted to scattering data \todo{check that this the case for the used values}
are shown in Table \ref{tab:apls}.
\todo{finalize figure (NMR headgr. OPs + SAXS, continue in the discussion.}
\begin{table}
  \caption{Area per lipid from different models for POPC without ions\label{tab:apls} }
  \begin{tabular}{c c c}
    model          & A (Å$^2$)   & Temperature [K] \\
    \hline
    Lipid14 (literature)  & 65.6$\pm$ 0.5  &  303 \\
    Lipid14ecc0.80+sigma0.875 &        &  313    \\
    GMX small patch           & 64.9   &         \\
    GMX 4xbig patch           & 65.5   &         \\
    oMM small patch           & 63.65  &         \\
    oMM 4xbig patch           & 63.7   &         \\
    \hline
    experiment \cite{Jambeck2012}\todoii{REF}{put original references, not Slipids param. paper.}  & 62.7  &  293    \\
    experiment  & 64.3  &  303    \\
    experiment  & 67.3  &  323    \\
    experiment  & 68.1  &  333    \\
    experiment POPE  & 56.6 &  303    \\
    \hline
  \end{tabular}
\end{table}

Ion binding in lipid bilayers can be accurately measured and compared
between experiments and simulations by using lipid headgroup order
parameters and ''electrometer concept'' introduced by Seelig et al. \cite{seelig87,catte16}.
This is based on the experimental observations that order parameters
for $\alpha$ and $\beta$ carbons are proportional to the amount of bound charge
in lipid bilayer, which was rationalized as a charge induced tilt
of the headgroup dipole \cite{seelig87}. Later analysis including signs of the order
parameters showed that the order parameters are actually decreasing with bound
positive charge and {\it vice versa} for negative charge \cite{ollila16,catte16}.
To date, there is no lipid model that can describe such a behaviour, 
the electrometer response and cation binding, quantitatively well~\cite{catte16}

@@

Order parameter changes from experiments \cite{seelig87} and simulations with Lipid14 model and with ECC-lipids model 
as a function of bound charge are shown in Fig. \ref{OrderParameterCHANGESnewMODELS}. 
Approximately linear decrease of headgroup order parameters is observed both 
in simulations and experiments with the cationic surfactant (dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$). 
The use of a cationic surfactant has the benefit that it is not a subject of partitioning between water and lipid phase. 
This means that we know the exact amount of bound charge in the membrane both in experiment and simulation. 
Quantitative comparison reveals that in the original Lipid14 model the response is overestimated for both order parameters, $\alpha$ and $\beta$. 
This suggests that the observed overestimated response in \cite{catte16} 
is at least in part due to the high sensitivity of the headgroup order parameter response, 
not only due to overestimated binding of cations. 
Secondly, the $S^\alpha/S^\beta$ ratio of the response is found to be slightly larger than in experiments (VALexp vs VALsim\todoi{Add values of $S^\alpha/S^\beta$ response for sim and experiment}).
This property is kept also by the newly derived model, ECC-lipid, and, hence, 
it shows a perfect agreement in the response of the headgroup order parameter $S^\alpha$ 
but underestimates slightly the response of $S^\beta$.

\todo{ongoing,Actual concentration of cations in simulation has yet to be estimated. 
If it varies too much from the nominal concentration, I may need to tweak the scaling factors, $f_q$ or only $f_\sigma$, to accomodate it. 
However, it is very unlikely, response to the surfactant \ce{DHMDMAB} is OK. 
Big patches with loads of solvent are running at the moment to guide me on the possible finite-size errors and this conc-error.}

>>>>

To date, there is no lipid model that can describe such a behaviour, 
the electrometer response and cation binding, quantitatively well~\cite{catte16}

Almost all lipid bilayer simulations reported in the literature 
are done with force fields that represent interactions between atoms
with pairwise additive empirical potential and exclude the electronic 
polarizability \cite{SOME_review, FF_papers}. While the issue has been 
considered higly relevant since the early days of lipid bilayer simulations,
practical solutions have turned out to be complicated \cite{??}.
Here we propose that electronic polarizability can be effectively included
in lipid bilayer simulations by using the electronic contimuun correction \cite{leontyev11}.

@@

\subsection{Electronic continuum correction for lipid bilayers}
According to the electronic continuum correction (ECC) the electronic
polarizability can be effectively included in MD simulations by scaling
the partial charges with a constant factor $f_q = \epsilon _{el} ^{-1/2}$, 
where $\epsilon _{el}$ is the electronic part of the dielectric constant of 
media \cite{leontyev11}. Measurements of high frequency 
dielectric constant gives values of approximately $\epsilon _{el} \approx 2$ 
for almost any biomaterial \cite{some_original_work, leontyev11}. % that exists in living cells. 
The value measured for water, $\epsilon _{el} = 1.78$, gives 
a scaling factor of $f_q = 0.75$ \cite{some_orig_source, leontyev11}, which has been
successfully used to improve the performance of ion force fields \cite{kohagen14,kohagen16,??}.

Here we apply the same approach to lipid bilayers to accurately 
describe the lipid headgroup response to Na+ and Ca2+ concentrations \cite{catte16}. 
We use the Lipid14~\cite{dickson14} force field parameters as a starting point,
because they give the most realistic headgroup response with added cations 
and relatively realistic glycerol backbone and headgroup structures
when compared with other state of the art lipid models \cite{botan15,catte16}.
The partial atomic charges in Lipid14 were derived by fitting the electrostatic 
potential to its model quantum chemistry representation in vacuum.
\todo{I do not fully understand the end of the paragraph} 
If such charges are obtained in an implicit solvent, they vary the most for the polar moieties \cite{maiejewski14}. 
In order to represent this difference in polarity in a model with fixed charges, 
we shall use the average charges from both environments (IpolQ charges) \cite{Cerutti2013}. 
By taking the charges of oxygen atoms in vacuum and implicit water solvent for POPC from \cite{maciejewski14}, 
we represent the IpolQ charges in the electronic continuum correction of the Lipid14 model
by increasing the scaling factor for charges to $f_q = 0.8$. 
The newly presented ECC-POPC model is based on the POPC model from

Hydrocarbon chains in Lipid14 and other lipids models are highly optimized and
give generally a good description for hydrophobic part of lipid bilayers
in various conditions \cite{ollila16},
in contrast to glycerol backbone and headgroup regions which require
some improvement in all available lipid models \cite{botan15}. 
To minimize the intercalation with the highly optimized hydrocarbon
parameters, we decided to apply ECC correction only
to the headgroup, glycerol backbone and carbonyl in acyl chains.
These regions are also most polar parts in lipids and are expected
to most affect on cation binding.   
%Because the biologically relevant cations like \ce{Na^+} and \ce{Ca^{2+}}
%are found to interact only with the phospholipid head group,
%and the accuracy of alkyl sidechains in Lipid14 model are satisfactory, 
%we exclude alkyl tails from our modifications.

The scaling of these partial charges by a factor of 0.8 reduce the 
area per molecule to ??, which is significantly smaller than the
experimental and original Lipid 14 values.  
The decrease of area was speculated 
\todo{Did you analyze the effect of this to hydration or did we only speculate?}
to arise from reduced hydration
of headgroup due to the lower polarity of molecules with scaled charges.
The hydration can be increased by decreasing the radius of atoms
by reducing $\sigma$ term in Lennart-Jones potential.
This allows water molecules approach closer to lipid atoms,
where electrostatic interactions are stronger and, thus, 
increases hydration level.
\todo{This effect may have an official name. In that case we should mention it.}
\todo{Is there some justification for the below sentence?} 
In line with the electronic continnum correction,
the scaling factor for sigma parameters, $f_\sigma$,
has to lay between $f_q$ and 1. Scaling factor
value of ?? for $\sigma$ was found to reproduce
area per molecule values close to experiments and
original Lipid14 model when charges were scaled with
0.8.

\subsection{Comparison to experimental data}
Structures sampled by individual lipid molecules were compared
to experimental data by using C-H bond order parameters \cite{ollila16}
\begin{equation}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between C-H bond and membrane
normal and average is taken over all sampled configurations.
Bilayer dimensions were compared to experiments by using the
scattering form factor \cite{ollila16}
\todo{Add equation}.

Ion binding was compared between experiments and simulations by 
using lipid headgroup order parameters and ''electrometer concept'' 
introduced by Seelig et al. \cite{seelig87,catte16}.
The concept is based on the experimental observation that the 
order parameters of $\alpha$ and $\beta$ carbons in lipid headgroup
(see Fig. \ref{??}) are proportional to the amount of bound charge
in lipid bilayer \cite{seelig87}. More recent analysis included also
the order parameter signs and concluded that the order parameters  
decrease with bound positive charge and increase with bound negative 
charge \cite{ollila16,catte16}. The observations are rationalized 
as a change of lipid headgroup dipole tilt to more vertical orientation
with bound positive charge and {\it vice versa} for negative \cite{seelig87}.

\todo{ongoing,Actual concentration of cations in simulation has yet to be estimated. 
If it varies too much from the nominal concentration, I may need to tweak the scaling 
factors, $f_q$ or only $f_\sigma$, to accomodate it. However, it is very unlikely, response 
to the surfactant \ce{DHMDMAB} is OK. Big patches with loads of solvent are running at the 
moment to guide me on the possible finite-size errors and this conc-error.
SAMULI: I thought about this a little bit. My current conclusion is that
we can take the number concentration from simulations with large box having 
a clear bulk concentration region and calculate the bulk concentration in mol/l
unit like this}
\begin{equation}
  C_{eq}[mol/l]=\frac{C_{plateau}[mM]}{0.602}
\end{equation}

\subsection{Simulation details}

\todo{To be written}

@@

\subsection{Lipid bilayers without ions}
The structural quality of
new ECC lipid model is evaluated against scattering form factors and NMR order parameters
in Fig. \ref{simVSexpNOions}
The optimal value for $f_\sigma$ was found by representing the overall membrane structure well
by matching scattering form factors to experimental data~\cite{Petrache06, Kucerka08, Pabst10}.
This ensures that the presented model describes correctly the membrane strucutre,
adopts a liquid disordered phase, 
and provides area per lipid agreeing with experiments (see Fig~\ref{simVSexpNOions} and Table~\ref{tab:apls}).

Order parameter changes from experiments \cite{seelig87} and simulations with Lipid14 model and with ECC-lipids model 
as a function of bound charge are shown in Fig. \ref{OrderParameterCHANGESnewMODELS}. 
Approximately linear decrease of headgroup order parameters is observed both 
in simulations and experiments with the cationic surfactant (dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$). 
The use of a cationic surfactant has the benefit that it is not a subject of partitioning between water and lipid phase. 
This means that we know the exact amount of bound charge in the membrane both in experiment and simulation. 
Quantitative comparison reveals that in the original Lipid14 model the response is overestimated for both order parameters, $\alpha$ and $\beta$. 
This suggests that the observed overestimated response in \cite{catte16} 
is at least in part due to the high sensitivity of the headgroup order parameter response, 
not only due to overestimated binding of cations. 
Secondly, the $S^\alpha/S^\beta$ ratio of the response is found to be slightly larger than in 
experiments (VALexp vs VALsim\todoi{Add values of $S^\alpha/S^\beta$ response for sim and experiment}).
This property is kept also by the newly derived model, ECC-lipid, and, hence, 
it shows a perfect agreement in the response of the headgroup order parameter $S^\alpha$ 
but underestimates slightly the response of $S^\beta$. 
\todo{Add acyl chain order parameters, POPC chemical structure}
\todo{Why original Lipid14 model seems to deviate from experiments, in contrast to the original publication? Different simulation setup used?}
\begin{figure}[]
  \centering
  \includegraphics[height=8.5cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
%  \caption{\label{fig:form-f}
%    X-ray scattering form factors from experiments~\cite{Kucerka2011} and simulations using Lipid14 and ECCLipids17 models.  }
%\end{figure}
%
%\begin{figure}[]
%  \centering
  \includegraphics[width=9.0cm]{../Fig/ipython_nb/Headgr_OPs_exp-L14-ECCL17.png}
  \caption{\label{simVSexpNOions}
    X-ray scattering form factors from experiments~\cite{Kucerka2011} and simulations using Lipid14 and ECCLipids17 models. 
    Headgroup and glycerol backbone order parameters from simulations with Lipid14 \cite{dickson14} and EECLipid17 models
    compared with experimental order parameters from \cite{ferreira13}.}
\end{figure}
Area per molecules extracted from MD simulations and SPD model fitted to scattering data \todo{check that this the case for the used values}
are shown in Table \ref{tab:apls}.
\todo{finalize figure (NMR headgr. OPs + SAXS, continue in the discussion.}
\begin{table}
  \caption{Area per lipid from different models for POPC without ions\label{tab:apls} }
  \begin{tabular}{c c c}
    model          & A (Å$^2$)   & Temperature [K] \\
    \hline
    Lipid14 (literature)  & 65.6$\pm$ 0.5  &  303 \\
    Lipid14ecc0.80+sigma0.875 &        &  313    \\
    GMX small patch           & 64.9   &         \\
    GMX 4xbig patch           & 65.5   &         \\
    oMM small patch           & 63.65  &         \\
    oMM 4xbig patch           & 63.7   &         \\
    \hline
    experiment \cite{Jambeck2012}\todoii{REF}{put original references, not Slipids param. paper.}  & 62.7  &  293    \\
    experiment  & 64.3  &  303    \\
    experiment  & 67.3  &  323    \\
    experiment  & 68.1  &  333    \\
    experiment POPE  & 56.6 &  303    \\
    \hline
  \end{tabular}
\end{table}

\subsection{Cation binding in POPC bilayer}
